---
title: "Data Analysis"
format: html
editor: visual
---

This document contains a descriptive and exploratory analysis of the data collected for the [Data Science for Open WASH data course](https://ds4owd-002.github.io/website/) registration form. 

```{r}
#| label: setup
#| include: false

# Set global options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# Load required libraries
library(tidyverse)
library(robotoolbox)
library(readxl)
library(labelled)

# Define OWD color palette
owd_palette <- c("#5b195b", "#9b2c60", "#ce525b",
                 "#f08453", "#ffbd54", "#f9f871")
```


```{r}
#| label: load-data

form_id <- "ashaEDvw4ZLwGi9bqXGeqb"
form_file <- "registration_form.xlsx"

# fetch data from Kobo with robotoolbox
raw_data <- kobo_data(x = form_id,
                      all_versions = TRUE)

# read in questionnaire and labels from XLS form 
questionnaire <- read_xlsx(form_file,
                           sheet = "survey")

label_dict <- read_xlsx("registration_form.xlsx",
                        sheet = "choices")
```

```{r}
#| label: data-processing

registration_data <- raw_data |> 
  select(github_username,
         gender,
         age_group,
         country_residence,
         education,
         employment_situation,
         org_type,
         prog_general,
         prog_r,
         prog_python,
         starts_with("llm_platforms")) |> 
  distinct(github_username, .keep_all = TRUE)

# starts_with("llm_") & !starts_with("llm_platforms")
```

## Categorical Variables Distribution

### Gender

```{r}
#| label: plot-gender

# Get gender labels
gender_labels <- label_dict |> 
  filter(list_name == "list_gender") |> 
  select(name, label)

plot_gender <- registration_data |> 
  count(gender) |> 
  left_join(gender_labels, by = c("gender" = "name")) |> 
  mutate(gender_label = coalesce(label, gender)) |> 
  ggplot(aes(x = reorder(gender_label, n), y = n)) +
  geom_col(fill = owd_palette[1]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Gender", y = "Count") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_gender)
```

### Age group

```{r}
#| label: plot-age

# Get age group labels
age_labels <- label_dict |> 
  filter(list_name == "list_age_group") |> 
  select(name, label)

plot_age <- registration_data |> 
  count(age_group) |> 
  left_join(age_labels, by = c("age_group" = "name")) |> 
  mutate(age_label = coalesce(label, age_group)) |> 
  ggplot(aes(x = reorder(age_label, n), y = n)) +
  geom_col(fill = owd_palette[2]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Age Group", y = "Count") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_age)
```

### Country of residence

```{r}
#| label: plot-country

# Get country labels
country_labels <- label_dict |> 
  filter(list_name == "countries") |> 
  select(name, label)

plot_country <- registration_data |> 
  count(country_residence) |> 
  left_join(country_labels, by = c("country_residence" = "name")) |> 
  mutate(country_name = coalesce(label, country_residence),
         rank = min_rank(desc(n))) |> 
  filter(rank <= 10) |>  # Show top 10 countries with ties
  ggplot(aes(x = reorder(country_name, n), y = n)) +
  geom_col(fill = owd_palette[3]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Country", y = "Count", 
       title = "Top 10 Countries") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_country)
```

### Education level

```{r}
#| label: plot-education

# Get education labels and remove text in parentheses
education_labels <- label_dict |> 
  filter(list_name == "education_level") |> 
  select(name, label) |> 
  mutate(label = str_remove(label, "\\s*\\(.*\\)"))

plot_education <- registration_data |> 
  count(education) |> 
  left_join(education_labels, by = c("education" = "name")) |> 
  mutate(education_label = coalesce(label, education)) |> 
  ggplot(aes(x = reorder(education_label, n), y = n)) +
  geom_col(fill = owd_palette[4]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Education Level", y = "Count") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_education)
```

### Employment situation

```{r}
#| label: plot-employment

# Get employment situation labels
employment_labels <- label_dict |> 
  filter(list_name == "employment_status") |> 
  select(name, label)

plot_employment <- registration_data |> 
  count(employment_situation) |> 
  left_join(employment_labels, by = c("employment_situation" = "name")) |> 
  mutate(employment_label = coalesce(label, employment_situation)) |> 
  ggplot(aes(x = reorder(employment_label, n), y = n)) +
  geom_col(fill = owd_palette[5]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Employment Situation", y = "Count") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_employment)
```

### Organization type

```{r}
#| label: plot-org-type

# Get organization type labels
org_labels <- label_dict |> 
  filter(list_name == "organisation_type") |> 
  select(name, label)

plot_org <- registration_data |> 
  count(org_type) |> 
  left_join(org_labels, by = c("org_type" = "name")) |> 
  mutate(org_label = coalesce(label, org_type)) |> 
  ggplot(aes(x = reorder(org_label, n), y = n)) +
  geom_col(fill = owd_palette[6]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "Organization Type", y = "Count") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_org)
```

### Programming skills

```{r}
#| label: plot-programming-skills
#| fig-height: 6

# Get programming level labels
prog_labels <- label_dict |> 
  filter(list_name == "programming_exp") |> 
  select(name, label)

# Combine programming experience variables
prog_data <- registration_data |> 
  select(prog_general, prog_r, prog_python) |> 
  pivot_longer(everything(), 
               names_to = "skill_type", 
               values_to = "level") |> 
  left_join(prog_labels, by = c("level" = "name")) |> 
  mutate(level_label = coalesce(label, level),
         skill_type = case_when(
           skill_type == "prog_general" ~ "General Programming",
           skill_type == "prog_r" ~ "R Programming",
           skill_type == "prog_python" ~ "Python Programming"
         ))

plot_programming <- prog_data |> 
  count(skill_type, level_label) |> 
  ggplot(aes(x = reorder(level_label, n), y = n, fill = skill_type)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(x = "Experience Level", y = "Count", fill = "Skill Type") +
  theme_minimal() +
  scale_fill_manual(values = owd_palette[1:3]) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_programming)
```

### LLM usage

```{r}
#| label: plot-llm-platforms
#| fig-height: 6

# Process LLM platforms data - convert all to character first
llm_platforms_data <- registration_data |> 
  select(starts_with("llm_platforms")) |> 
  mutate(across(everything(), as.character)) |> 
  mutate(id = row_number()) |> 
  pivot_longer(-id, 
               names_to = "platform_var", 
               values_to = "platform") |> 
  filter(!is.na(platform) & platform != "NA") |> 
  separate_rows(platform, sep = " ") |> 
  filter(platform != "" & platform != "0" & platform != "1")

# Get platform labels if available
platform_labels <- label_dict |> 
  filter(list_name == "llm_tools") |> 
  select(name, label)

# Count and plot platform usage - top 10 with ties
plot_llm_platforms <- llm_platforms_data |> 
  count(platform) |> 
  left_join(platform_labels, by = c("platform" = "name")) |> 
  mutate(platform_label = coalesce(label, platform),
         rank = min_rank(desc(n))) |> 
  filter(rank <= 10) |>  # This will include ties at rank 10
  ggplot(aes(x = reorder(platform_label, n), y = n)) +
  geom_col(fill = owd_palette[2]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(x = "LLM Platform", y = "Number of Users",
       title = "Top 10 LLM Platforms Used by Participants") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

print(plot_llm_platforms)
```

## Storytelling

```{r}
#| label: data-facts
#| include: false

# Total registrations
total_registrations <- nrow(registration_data)

# Top countries
country_counts <- registration_data |> 
  count(country_residence) |> 
  left_join(label_dict |> 
            filter(list_name == "countries") |> 
            select(name, label), 
            by = c("country_residence" = "name")) |> 
  mutate(country_name = coalesce(label, country_residence)) |> 
  arrange(desc(n))

top1_country <- country_counts$country_name[1]
top1_count <- country_counts$n[1]
top2_country <- country_counts$country_name[2]
top2_count <- country_counts$n[2]
top3_country <- country_counts$country_name[3]
top3_count <- country_counts$n[3]

# Gender diversity
gender_counts <- registration_data |> 
  count(gender) |> 
  mutate(percent = round(n/sum(n) * 100))

female_percent <- gender_counts |> 
  filter(gender == "woman") |> 
  pull(percent)

# Programming experience
prog_r_experience <- registration_data |> 
  count(prog_r) |> 
  left_join(label_dict |> 
            filter(list_name == "programming_exp") |> 
            select(name, label), 
            by = c("prog_r" = "name")) |> 
  mutate(experience_level = coalesce(label, prog_r),
         percent = round(n/sum(n) * 100))

r_beginners <- prog_r_experience |> 
  filter(prog_r %in% c("none", "beginner")) |> 
  summarise(total_percent = sum(percent)) |> 
  pull(total_percent)

# Organization diversity
org_diversity <- registration_data |> 
  count(org_type) |> 
  nrow()

# Age diversity
age_range <- registration_data |> 
  count(age_group) |> 
  left_join(label_dict |> 
            filter(list_name == "list_age_group") |> 
            select(name, label), 
            by = c("age_group" = "name")) |> 
  mutate(age_label = coalesce(label, age_group))

most_common_age <- age_range |> 
  slice_max(n) |> 
  pull(age_label)

# Education levels
education_high <- registration_data |> 
  count(education) |> 
  mutate(percent = round(n/sum(n) * 100)) |> 
  filter(education %in% c("master", "phd")) |> 
  summarise(total = sum(percent)) |> 
  pull(total)

# LLM usage stats
llm_platforms_summary <- registration_data |> 
  select(starts_with("llm_platforms")) |> 
  mutate(across(everything(), as.character)) |> 
  mutate(id = row_number()) |> 
  pivot_longer(-id, 
               names_to = "platform_var", 
               values_to = "platform") |> 
  filter(!is.na(platform) & platform != "NA") |> 
  separate_rows(platform, sep = " ") |> 
  filter(platform != "" & platform != "0" & platform != "1") |> 
  count(platform) |> 
  left_join(label_dict |> 
            filter(list_name == "llm_tools") |> 
            select(name, label), 
            by = c("platform" = "name")) |> 
  mutate(platform_label = coalesce(label, platform),
         platform_label_clean = str_remove(platform_label, "\\s*\\(.*\\)")) |> 
  arrange(desc(n))

top_llm_tools <- paste(head(llm_platforms_summary$platform_label_clean, 3), collapse = ", ")
llm_users_percent <- round(sum(!is.na(registration_data$llm_platforms)) / nrow(registration_data) * 100)
```

ðŸŒ **Data Science for openwashdata: Building a Global Community!** 

We're thrilled to announce **`r total_registrations` registrations** from passionate learners worldwide who are ready to transform water, sanitation, and hygiene data into actionable insights! ðŸ’§ðŸ“Š

Registrations are still open: https://ee-eu.kobotoolbox.org/single/7V3qeDYD

**ðŸ—ºï¸ Global reach, diverse perspectives:** Our participants span continents, with `r top1_country` leading the way (`r top1_count` ), followed by strong representation from `r top2_country` (`r top2_count`) and `r top3_country` (`r top3_count`). We're happy to see a strong female participation `r female_percent`%. You're creating an inclusive learning environment where diverse voices will enrich our experience. ðŸ¤

**ðŸ“š Meeting learners at every level:** While `r education_high`% of participants hold advanced degrees, `r r_beginners`% are beginners in R programming. This mix creates a unique learning environment where everyone brings valuable perspectives - whether you're taking your first steps in coding or bringing years of WASH expertise ready to add data science to your toolkit. ðŸ’»âœ¨

**ðŸ¤– AI-powered learning:** With `r llm_users_percent`% of participants already using AI tools for coding (most popular: `r top_llm_tools`), we're excited to include our bonus module: **"Use of AI for coding support"**. This module will help all participants - from beginners to experienced coders - leverage AI tools effectively and responsibly in their data science journey.

Join this vibrant community committed to using data science for open research and data. Together, we're building the knowledge to tackle water and sanitation challenges. ðŸ’ªðŸŒŠ

#DataScience #WASH #OpenData #OpenScience #RStats 
